### 2023-11-01 journal/boundedspace.tex

\section{Bounding Space Usage}
\label{reducing}

Operations remain in the \fld{blocks} arrays forever. 
This uses space proportional to the number of enqueues that have been invoked.
Now, we modify the implementation to remove blocks that are no longer needed, so that space usage is
polynomial in $p$ and $q$, while (amortized) step complexity is still polylogarithmic. We replace the \fld{blocks} array in each node by a red-black tree (RBT)
that stores the blocks.
Each block has an additional \fld{index} field
that represents its position within the original \fld{blocks} array, and
blocks in a RBT are sorted by \fld{index}.
The attempt to install a new block in $\fld{blocks}[i]$  on line \ref{cas}
is replaced by an attempt to insert a new block with index $i$ into the RBT.
Accessing the block in $\fld{blocks}[i]$ is 
replaced by searching the RBT for the  index $i$.
The binary searches for a block in line \ref{FRb} and \ref{getChild} can simply search the RBT
using the \fld{sum\sub{enq}} field, since the RBT is also sorted with respect to this field, by Invariant \ref{inv::sum}.

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree operations.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a block, just as the \op{CAS} on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful \op{CAS}.
Moreover, the insertion should succeed only if the newly inserted block has a larger index than any other block in the RBT.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all RBT nodes are immutable, and operations on the 
RBT make a new copy of each RBT node $x$ that must be modified, as well
as each RBT node along the path from the RBT's root to~$x$.
The RBT reachable from the new copy of the root is the result of applying the RBT operation.
This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
Once a process has performed an update to the RBT representing the blocks of a node 
$\var{v}$ in the \ordering\ tree, 
it uses a CAS to swing $\var{v}$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent RBTs in a similar way for a snapshot data structure.

To prevent RBTs from growing without bound, we would like to discard
blocks that are no longer needed.
Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
also keep the running time of our operations polylogarithmic.
Blocks should be kept if they contain operations still in progress.
Moreover, a block containing an \op{Enqueue}($x$) must be kept until $x$ is dequeued.

@@

Our GC scheme ensures each RBT has  $O(q+p^2\log p)$ blocks, so RBT operations take $O(\log(p+q))$ time.
Excluding GC, an operation does $O(1)$ operations on RBTs at each level of the tree for a total of
$O(\log p\log(p + q))$ steps.
A GC phase takes $O(p\log p \log(p+q))$ steps to help complete all pending operations.
If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(G\log(p+q))$ steps.
Since there are at least $G$ operations between two GC phases, each node contributes $O(\log(p+q))$ steps to each operation in an amortized sense.
Adding up over all nodes an operation may have to do GC on, 
an operation spends $O(\log p\log(p+q))$ steps doing GC in an amortized sense.
So the total amortized step complexity is  $O(\log p \log(p+q))$ per operation.
%Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

>>>>

\section{Bounded-Space Implementation}
\label{reducing}

In the implementation described in Section \ref{DescriptQ}, 
operations remain in the \fld{blocks} arrays forever. 
Thus, the space used continues to grow as operations are invoked.
Now, we modify the implementation to remove blocks that are no longer needed, so that space usage is
polynomial in $p$ and $q$, while ensuring the (amortized) step complexity is still polylogarithmic. We replace the \fld{blocks} array in each node by a red-black tree (RBT)
that stores the blocks.
Each block has an additional \fld{index} field
that represents its position within the original \fld{blocks} array, and
blocks in a RBT are sorted by \fld{index}.
The attempt to install a new block in $\fld{blocks}[i]$  on line \ref{cas}
is replaced by an attempt to insert a new block with index $i$ into the RBT.
Accessing the block in $\fld{blocks}[i]$ is 
replaced by searching the RBT for the  index $i$.
The binary searches for a block in line \ref{FRb} and \ref{getChild} can simply search the RBT
using the \fld{sum\sub{enq}} field, since the RBT is also sorted with respect to this field, by Invariant \ref{inv::sum}.

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree operations.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a block, just as the \op{CAS} on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful \op{CAS}.
Moreover, the insertion should succeed only if the newly inserted block has a larger index than any other block in the RBT.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all RBT nodes are immutable, and operations on the 
RBT make a new copy of each RBT node $x$ that must be modified, as well
as each RBT node along the path from the RBT's root to~$x$.
The RBT reachable from the new copy of the root is the result of applying the RBT operation.
This  adds only a constant factor to the running time of any routine designed for a (sequential) RBT.
Once a process has performed an update to the RBT representing the blocks of a node 
$\var{v}$ in the \ordering\ tree, 
it uses a CAS to swing $\var{v}$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent RBTs in a similar way for a snapshot data structure.

To prevent RBTs from growing without bound, we must discard
blocks that are no longer needed.
Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
also keep the running time of our operations polylogarithmic.
Blocks should be kept if they contain operations still in progress.
Moreover, a block containing an \op{Enqueue}($x$) operation must be kept until $x$ is dequeued.

@@

%Our GC scheme ensures each RBT has  $O(q+p^2\log p)$ blocks, so RBT operations take $O(\log(p+q))$ time.
%Excluding GC, an operation does $O(1)$ operations on RBTs at each level of the tree for a total of
%$O(\log p\log(p + q))$ steps.
%A GC phase takes $O(p\log p \log(p+q))$ steps to help complete all pending operations.
%If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(G\log(p+q))$ steps.
%Since there are at least $G$ operations between two GC phases, each node contributes $O(\log(p+q))$ steps to each operation in an amortized sense.
%Adding up over all nodes an operation may have to do GC on, 
%an operation spends $O(\log p\log(p+q))$ steps doing GC in an amortized sense.
%So the total amortized step complexity is  $O(\log p \log(p+q))$ per operation.
%Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

### 2023-11-01 journal/correctness.tex

\er{edited this far}

We first show the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.  This is proved by showing that the \fld{super} field of a block $B$ in node $\var{v}$ is read from  
\var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in the parent node.
On one hand, $B.super$ is written before $B_s$ is installed:
\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
$B.\fld{super}$ is written after $B$ is installed, and Lemma \ref{successfulRefresh} ensures that $B$ is propagated to the
parent soon after.

\begin{lemma}
\label{superRelation}
Let $B=\var{v}.\var{blocks}[b]$.
If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{lemma}
\begin{proof}
We first show that $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
in $\var{v}.\fld{parent.blocks}[s]$.
By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
By Invariant \ref{inv::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by Invariant \ref{inv::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by Observation \ref{nonDecreasingHead}, it follows that $B.super\leq s$.

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least 1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
By Lemma \ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
$\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $\var{v}.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $\var{v}$;
otherwise, Lemma \ref{successfulRefresh} would imply that $B$ is a subblock of one of 
$\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($\var{v}, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $\var{v}.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $\var{v}.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By Observation \ref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $\var{v}.\fld{parent}.\head$.
\end{proof}

%The reader may wonder when the case $b\nf{.super}=s$ happens. This can happen when $
%\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block 
%into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
To show  \op{GetEnqueue} and \op{IndexDequeue} work correctly, we  just  check that they correctly compute the index of the required block
and the operation's rank within the block.  
For \op{IndexDequeue}, we use Lemma \ref{superRelation} each time \op{IndexDequeue} goes one step up the tree.
\here{mention somewhere why precondition of IndexDequeue is true? also check that block of parent indexed by sup in that routine is non-null}

@@

Next, the position of the required dequeue in $D(B')$ is computed in 
lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
We first add the number of dequeues in the subblocks of $B'$ in $\var{v}$ that precede $B$ on line \ref{computeISuperStart}.
If $\var{v}$ is the right child of its parent, then all of the subblocks of $B'$ from $\var{v}$'s left sibling
also precede the required dequeue, so we add the number of dequeues in those subblocks in line \ref{considerLeftBeforeRight}.

@@

Assuming the claim holds for $\var{v}$'s children, we prove it for $\var{v}$.
Let $B$ be $\var{v}.\fld{blocks}[b]$.
By (\ref{defSeqs}),
$E(B)$ is obtained by concatenating the enqueue sequences of the direct subblocks
of $B$, which are listed in (\ref{defsubblock}).
By Invariant \ref{inv::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
of enqueues in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required enqueue.
Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
By Invariant \ref{inv::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
enqueues in $E'$ before the enqueues of block $B'$, so
the value $i'$ computed on line \ref{getChildIndex} is the position of the required enqueue within $E(B')$.
Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
returns the required result, by the induction hypothesis.
\end{proof}

@@

We show that the linearization ordering $L$ defined in 
(\ref{linearization}) is a legal permutation of a subset of the operations in 
the execution, i.e., that it includes all operations that terminate and 
if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.  We also show the result each dequeue returns is the same as in the sequential execution  $L$.

@@

\begin{lemma}\label{linearCorrect}
Each terminating dequeue returns the response it would in the sequential execution~$L$.
\end{lemma}
\begin{proof}
If a dequeue $D$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
By Lemma \ref{lem::indexDequeue}, $D$'s call to \op{IndexDequeue} on line \ref{invokeIndexDequeue}
returns a pair $\langle b,i\rangle$ such that $D$ is the $i$th dequeue in the block 
$B=\var{root}.\fld{blocks}[b]$.
$D$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
By Lemma \ref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
in the order given by $L$.
By Invariant \ref{inv::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
is the number of enqueues in $B$.
Since the enqueues in block $B$ precede the dequeues,
the queue is empty when the $i$th dequeue of $B$ occurs if 
$\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
So $D$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
execution $L$.
Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
in the sequential execution $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
the number of non-\nl\ dequeues in that prefix of $L$.
Hence, line \ref{computeE} sets $e$ to the rank of $D$ among all the non-\nl\ dequeues in $L$.
Thus, in the sequential execution~$L$, $D$ returns the value enqueued by the $e$th enqueue in $L$.
By Invariant \ref{inv::sum}, this enqueue is the $i_e$th enqueue 
in $E(\var{root}.\fld{blocks}[b_e])$, where
$b_e$ and $i_e$ are the values $D$ computes on line \ref{FRb} and \ref{FRi}.
By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
\end{proof}

@@

\here{Should there be a lemma somewhere in this section that explicitly shows that each
object we dereference is non-null?}

>>>>

In this section, we show that the routines used to compute responses to \op{Dequeue}
operations work correctly.
We first prove the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.  We do this by showing that the \fld{super} field of a block $B$ in node $\var{v}$ is read from  
\var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in $v$'s parent.
In particular, $B.super$ is written after $B$ is installed, but before $B_s$ is installed.
%On one hand, $B.super$ is written before $B_s$ is installed:
%\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
%$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
%On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
%$B.\fld{super}$ is written after $B$ is installed, and Lemma \ref{successfulRefresh} ensures that $B$ is propagated to the
%parent soon after.

\begin{lemma}
\label{superRelation}
Let $B=\var{v}.\var{blocks}[b]$.
If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{lemma}
\begin{proof}
We first show that $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
in $\var{v}.\fld{parent.blocks}[s]$.
By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} %or \ref{createEndRight} 
obtains a value greater than $b$.
By Invariant \ref{inv::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by Invariant \ref{inv::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by Observation \ref{nonDecreasingHead}, it follows that $B.super\leq s$.

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least~1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
By Lemma~\ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
$\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $\var{v}.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $\var{v}$;
otherwise, Lemma \ref{successfulRefresh} would imply that $B$ is a subblock of one of 
$\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($\var{v}, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $\var{v}.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $\var{v}.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By Observation \ref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $\var{v}.\fld{parent}.\head$.
\end{proof}

%The reader may wonder when the case $b\nf{.super}=s$ happens. This can happen when $
%\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block 
%into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
To show  \op{GetEnqueue} and \op{IndexDequeue} work correctly, we  just  check that they correctly compute the index of the required block
and the operation's rank within the block.  
For \op{IndexDequeue}, we use Lemma \ref{superRelation} each time \op{IndexDequeue} goes one step up the tree.
%\here{mention somewhere why precondition of IndexDequeue is true? also check that block of parent indexed by sup in that routine is non-null}

@@

Next, the position of the required dequeue in $D(B')$ (as defined by Equation (\ref{defSeqs})) is computed in 
lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
We first add the number of dequeues in the subblocks of $B'$ in $\var{v}$ that precede $B$ on line~\ref{computeISuperStart}.
If $\var{v}$ is the right child of its parent, then all of the subblocks of $B'$ from $\var{v}$'s left sibling
also precede the required dequeue, so we add the number of dequeues in those subblocks in line \ref{considerLeftBeforeRight}.

@@

Assuming the claim holds for $\var{v}$'s children, we prove it for $\var{v}$.
Let $B$ be $\var{v}.\fld{blocks}[b]$.
By Equation (\ref{defSeqs}),
$E(B)$ is obtained by concatenating the enqueue sequences of the direct subblocks
of $B$, which are listed in (\ref{defsubblock}).
By Invariant \ref{inv::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
of enqueues in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required enqueue operation.
Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
By Invariant \ref{inv::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
enqueues in $E'$ before the enqueues of block $B'$, so
the value $i'$ computed on line \ref{getChildIndex} is the position of the required enqueue within $E(B')$.
Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
returns the required result, by the induction hypothesis.
\end{proof}

@@

We show that the linearization ordering $L$ defined in Equation
(\ref{linearization}) is a legal permutation of a subset of the operations in 
the execution, i.e., that it includes all operations that terminate and 
if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.  We also show each completed dequeue returns the same result as it would in the sequential execution~$L$.

@@

\begin{lemma}\label{linearCorrect}
Each terminating dequeue returns the response it would in the sequential execution~$L$.
\end{lemma}
\begin{proof}
If a dequeue $Deq$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
By Lemma \ref{lem::indexDequeue}, $Deq$'s call to \op{IndexDequeue} on line~\ref{invokeIndexDequeue}
returns a pair $\langle b,i\rangle$ such that $Deq$ is the $i$th dequeue in the block 
$B=\var{root}.\fld{blocks}[b]$.
$Deq$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
By Lemma~\ref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
in the order given by $L$.
By Invariant \ref{inv::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
is the number of enqueues in $B$.
Since the enqueues in block $B$ precede the dequeues,
the queue is empty when the $i$th dequeue of $B$ occurs if 
$\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
So $Deq$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
execution $L$.
Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
in the sequential execution $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
the number of non-\nl\ dequeues in that prefix of $L$.
Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ dequeues in $L$.
Thus, in the sequential execution~$L$, $Deq$ returns the value enqueued by the $e$th enqueue in $L$.
By Invariant \ref{inv::sum}, this enqueue is the $i_e$th enqueue 
in $E(\var{root}.\fld{blocks}[b_e])$, where
$b_e$ and $i_e$ are the values $Deq$ computes on line \ref{FRb} and \ref{FRi}.
By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
\end{proof}

@@

%\here{Should there be a lemma somewhere in this section that explicitly shows that each
%object we dereference is non-null?}
